Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

## Dimensionality reduction: ZINB-WaVE

@@

```{r boxplotNorm, fig.cap="Boxplot of normalized values. Different colors are for different batches."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = 'n')
abline(h=0)
```

@@

```{r pcanorm, fig.cap="PCA of normalized values where each point represent a cell. In the left panel, cells have been colored according to their batch. In the right panel, colors correspond to the original published clusters."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = '')
plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
```

@@

```{r mdsW, fig.cap="MDS of the low dimensional matrix `W` where each point represent a cell. Cells have been colored according to the original published clusters."}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(clus.labels)], main = '',
     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

The next step of the workflow is to cluster the cells according to the low dimensional matrix `W` computed in the previous step. We used function `RSEC` from bioconductor package clusterExperiment. Function `RSEC` implements a resampling-based sequential ensemble clustering method for identifying stable and tight cell clusters. The method aggregates multiple clustering results obtained from different base clustering algorithms and applications of a given algorithm to resampled versions of the learning set.

@@

We can visualize the resulting clusterings using the `plotClusters` function. See Figure \@ref(fig:examineCombineMany) where cells are in the columns, and different clusterings on the rows. Each sample is color coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap. The first raw correspond to a consensus clustering across the many different clusterings performed.

```{r examineCombineMany, fig.cap="Clusters found using function RSEC from clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

We can also visualize the proportion of times the consensus clusters were together across the different clusterings. See Figure \@ref(fig:plotcoclust).

```{r plotcoclust, fig.cap="Heatmap of the proportion of times the clusters were together across the different clusterings."}
plotCoClustering(ceObj)
```

Over stable and robust clustering, the number of cells in each cluster is as follow

@@

We can now plot a heatmap of the normalized values for the 1000 most variable genes where the cells have been clustered using the `RSEC` function. See Figure \@ref(fig:heatmapsClusters).

```{r heatmapsClusters, fig.cap="Heatmap of the normalized values for the 1000 most variable genes where genes are in the rows and cells in the columns."}
# set colors for cell clusterings
colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "clusterLabels" = origClusterColors)

@@

Finally, we can visualize the cells in a two dimensional space by plotting the MDS of the low dimensional matrix `W` coloring the cells by their newly found clusters. See Figure \@ref(fig:mdsWce) which is the same as Figure \@ref(fig:mdsW) but where colors are for the newly found clusters instead of the clusters found in the orginal published work.

```{r mdsWce, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the clusters found using our workflow."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = '', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

@@

In an effort to improve the workflow and analysis of scRNA-seq data, we are currently exploring statistical methods to perform DE analysis using the models we developed. One the one hand, we are developing a DE analysis method using the framework of slingshot which would take advantage of the lineages found. On the other hand, we are working on a DE analysis method using the ZINB-WaVE model which would account for the dropouts and gene and cell-level covariates. Additionally, we are working on a statistical method using the ZINB-WaVE model to impute the dropouts in the scRNA-seq data. This matrix with imputed counts could for example be used in the workflow in the dimensionality reduction, clustering or pseudo ordering steps.

Finally, a new R class called SingleCellExperiment should be released soon. This new class would essentially be a SummarizedExperiment class with a couple of additional slots, the most important of which is reducedDims, which, much like the assays slot of SummarizedExperiment can contain one or more matrices with reduced dimensional matrices. This new SingleCellExperiment class would be a valuable addition to the workflow as we could store in one object the raw counts as well as the low dimensional matrix created by the dimensionality reduction step. Once this class is implemented, we would like to incorporate it to the workflow.

@@

This workflow provides guidance to perform downstream analysis of scRNA-seq datasets in R. We propose a worflow in four steps: dimensionality reduction adjusting for cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the lineages. The workflow is general and flexible allowing the user to sustitute the statistical method used in each step of the worflow by a different method. We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help focusing on discovering new biological insights.

>>>>

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; and (3) inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step. [SD: Include DE as 4th step, to match statement in "Conclusion"?]

@@

## Normalization and dimensionality reduction: ZINB-WaVE

@@

```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures, color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = 'n')
abline(h=0)
```

@@

```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell. In the left panel, cells are color-coded by batch. In the right panel, cells are color-coded by original published clusters."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = '')
plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
```

@@

```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clusters."}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(clus.labels)], main = '',
     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[SD: Add ref.]. A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings. Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap. The first row correspond to a consensus clustering across all candidate clusterings.

```{r examineCombineMany, fig.cap="Clusters found using the function RSEC from the clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```
The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).

```{r plotcoclust, fig.cap="Heatmap of co-clustering matrix."}
plotCoClustering(ceObj)
```

The distribution of cells across the consensus clusters is as follows:

@@

Figure \@ref(fig:heatmapsClusters) displays a heatmap of the normalized expression measures for the 1,000 most variable genes, where cells are clustered according to the RSEC consensus.

```{r heatmapsClusters, fig.cap="Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "clusterLabels" = origClusterColors)

@@

Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is anologous to Figure \@ref(fig:mdsW) for the original published clusters.

```{r mdsWce, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = '', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

@@

In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model. In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference. ZINB-WaVE can also be used for identifying genes that are differentially expressed between cells, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns. We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.

Finally, a new R class called `SingleCellExperiment` should be released soon. This new class would essentially be a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment` can contain one or more matrices of reduced dimension. This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step. Once this class is implemented, we would like to incorporate it to the workflow.

@@

This workflow provides a tutorial for the downstream analysis of scRNA-seq data in R. The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages. The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method. We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help with the discovery of novel biological insights.
